Galactic Chemical Evolution Models Favor an Extended Type Ia Supernova Delay-Time Distribution

### 2024-02-08 src/tex/ms.tex
[showyourwork] overleaf sync

\paragraph{Two-infall} 
We parameterize the infall rate as two successive, exponentially declining bursts as in \citet{Chiappini2001-AbundanceGradients}:
\begin{equation}
    \label{eq:twoinfall-ifr}
    f_{\rm TI}(t|R_{\rm gal}) = N_1(R_{\rm gal}) e^{-t/\tau_1} + N_2(R_{\rm gal}) e^{-(t-t_{\rm on})/\tau_2}
\end{equation}
In this model, the first infall produces the thick disk and the second infall produces the thin disk. The normalization ratio $N_2/N_1$ is calculated so that the thick-to-thin-disk surface density ratio $f_\Sigma(R)=\Sigma_2(R)/\Sigma_1(R)$ is given by
\begin{equation}
    f_\Sigma(R) = f_\Sigma(0) e^{R(1/R_2 - 1/R_1)}.
\end{equation}
Following \citet{BlandHawthornGerhard2016-MilkyWayReview}, we adopt values for the thick disk scale radius $R_1=2.0$ kpc, thin disk scale radius $R_2=2.5$ kpc, and $f_\Sigma(0)=0.27$.

@@

The right-hand panel of Figure \ref{fig:onezone-twopanel} compares the one-zone model outputs from the full range of DTD models we investigate in this work. The choice of DTD primarily affects the location of the high-$\alpha$ ``knee'' in the [Fe/H]--[O/Fe] abundance tracks. At one extreme is the triple-system model, which sees the CC SN plateau extend up to [Fe/H]$\approx0.8$ followed by a sharp downward turn as the SN Ia rate suddenly increases at a delay time of 500 Myr. Star formation proceeds for such a long time before the knee that the [O/Fe] distribution shows a slight second peak around the CC SN value; this model is the only one which produces any amount of bimodality in the one-zone simulations. At the other extreme are the two-population and power-law ($\alpha=-1.1$) models which peak immediately after the minimum delay time of 40 Myr, placing the knee at [Fe/H]$\approx-1.8$. The two-population model has a unique second knee at [Fe/H]$\sim-0.2$ and [O/Fe]$\sim0.1$, which is a result of the delayed exponential component. The abundance tracks from the plateau ($W=1$ Gyr) and exponential ($\tau=1.5$ Gyr) models lie in the intermediate space between these extremes. On the other hand, the [O/Fe] distribution from the exponential model is the most strongly peaked at the Solar value, while the plateau ($W=1$ Gyr) and triple-system models produce nearly identical distributions in the low-$\alpha$ regime. The exponential ($\tau=3$ Gyr) and plateau ($W=0.3$ Gyr) models, while not shown, produce similar abundance tracks to the plateau ($W=1$ Gyr) and exponential ($\tau=1.5$ Gyr) models, respectively.

@@

While trends with $R_{\rm gal}$ and $|z|$ are similar, the shape of the distribution varies greatly with the chosen SFH. The inside-out and late-burst models produce similar distributions because of the similarity of their underlying SFHs \textemdash both skew heavily toward near-Solar [O/Fe], although the late-burst model produces a slightly broader peak and a less-prominent high-[O/Fe] tail. The only region which shows any significant skew toward high [O/Fe] is $R_{\rm gal}=3-5$ kpc and $|z|=1-2$ kpc, but the shift to higher [O/Fe] at high latitudes is gradual and does not produce the notable trough at [O/Fe]$\approx0.2$ which is seen in the APOGEE data.

@@

The two-infall SFH model produces \textit{three} distinct modes at [O/Fe]$\approx 0.0$, $0.2$, and $0.4$. At small $R_{\rm gal}$ and with increasing $|z|$, the low-$\alpha$ peak decreases in prominence and the high-$\alpha$ peak increases, but the intermediate peak is a striking feature at all latitudes which is at odds with the APOGEE data.

@@

Figure \ref{fig:ofe-df-dtd} shows [O/Fe] distributions produced by simulations with the same SFH but a range of different DTDs. We choose the early-burst SFH because it produces distinct high- and low-$\alpha$ sequences, which makes the effect of the DTD on the former more apparent. The most apparent effect of the DTD is to shift the mode of the high-$\alpha$ sequence. The two-population DTD, which has the most prompt SNe Ia of any model in this paper, places the high-$\alpha$ sequence at [O/Fe]$\approx 0.2$, while the triple-system DTD, which has the fewest prompt SNe Ia, places it $\sim0.2$ dex higher at [O/Fe]$\approx0.4$. This metric alone would seem to favor the exponential ($\tau=1.5$ Gyr) or plateau ($W=0.3$ Gyr; not shown) models; however,

@@

The inside-out and late-burst models produce similar results, the main difference being a slightly broader core of the abundance distribution around [O/Fe]$\approx0.05$ dex. The Gaussian burst in SFH in the late-burst model introduces a loop in the ISM abundance track, where an uptick in star formation at $t\approx11$ Gyr raises the CCSN rate, leading to a slight increase in [O/Fe] before the subsequent increase in the SN Ia rate brings it back down. This same pattern is seen much more strongly in the abundance tracks for the two-infall model: here, the significant infall of pristine gas at $t=4$ Gyr leads to rapid dilution of the ISM, followed by a large burst in the SFR which raises [O/Fe] by $\sim 0.2$ dex. In this case the second infall produces a ``ridge line'' in the abundance distribution at [O/Fe]$\approx 0.2$. Finally, the early-burst model shows the clearest separation between a high-[O/Fe] track which follows the ISM abundance from low metallicities, and a low-[O/Fe] sequence composed of stars with a wide range of birth radii. The early-burst model sees a higher density of points which closely follow the ISM track at [Fe/H]$\lesssim 0.5$ which is likely the result of an increased SFR at early times relative to the other models.

@@

For the early-burst SFH model, the simulated stellar ages are almost perfectly aligned with the data for [O/Fe]$\gtrsim 0.2$. The rapid rise in the SFE at early times delays the descent to lower [O/Fe] values and produces a clump of low-metallicity, high-[O/Fe] stars at an age of $\sim10$ Gyr. The early-burst model produces a sharp transition between a linear relationship between [O/Fe] and age for stars older than $\sim5$ Gyr and a flat relationship for the younger stars, and it too misses the lowest [O/Fe] bin in the data. Finally, the two-infall SFH model produces the closest overall agreement between the simulated and observed median ages. Stars with [O/Fe]$\gtrsim 0.25$ were produced in the first infall, while the second infall produces a clump of stars with similar metallicity, ages of $\sim8$ Gyr, and [O/Fe]$\approx0.2$. There is an interesting small population of very old, low-$\alpha$ stars which is not seen in any of the other models. The two-infall model is the only one which plateaus at sub-Solar [O/Fe] for the youngest stars, leading to a very good match to the observed data at low [O/Fe].

>>>>

\paragraph{Two-infall} [Expand]
We parameterize the infall rate as two successive, exponentially declining bursts as in \citet{Chiappini2001-AbundanceGradients}:
\begin{equation}
    \label{eq:twoinfall-ifr}
    f_{\rm TI}(t|R_{\rm gal}) = N_1(R_{\rm gal}) e^{-t/\tau_1} + N_2(R_{\rm gal}) e^{-(t-t_{\rm on})/\tau_2}
\end{equation}
In this model, the first infall produces the thick disk and the second infall produces the thin disk. The normalization ratio $N_2/N_1$ is calculated so that the thick-to-thin-disk surface density ratio $f_\Sigma(R)=\Sigma_2(R)/\Sigma_1(R)$ is given by
\begin{equation}
    f_\Sigma(R) = f_\Sigma(0) e^{R(1/R_2 - 1/R_1)}.
\end{equation}
Following \citet{BlandHawthornGerhard2016-MilkyWayReview}, we adopt values for the thick disk scale radius $R_1=2.0$ kpc, thin disk scale radius $R_2=2.5$ kpc, and $f_\Sigma(0)=0.27$.

@@

The right-hand panel of Figure \ref{fig:onezone-twopanel} compares the one-zone model outputs from the full range of DTD models we investigate in this work. The choice of DTD primarily affects the location of the high-$\alpha$ ``knee'' in the [Fe/H]--[O/Fe] abundance tracks. At one extreme is the triple-system model, which sees the CC SN plateau extend up to ${\rm [Fe/H]}\approx0.8$ followed by a sharp downward turn as the SN Ia rate suddenly increases at a delay time of 500 Myr. Star formation proceeds for such a long time before the knee that the [O/Fe] distribution shows a slight second peak around the CC SN value; this model is the only one which produces any amount of bimodality in the one-zone simulations. At the other extreme are the two-population and power-law ($\alpha=-1.1$) models which peak immediately after the minimum delay time of 40 Myr, placing the knee at ${\rm [Fe/H]}\approx-1.8$. The two-population model has a unique second knee at ${\rm [Fe/H]}\approx-0.2$ and ${\rm [O/Fe]}\approx0.1$, which is a result of the delayed exponential component. The abundance tracks from the plateau ($W=1$ Gyr) and exponential ($\tau=1.5$ Gyr) models lie in the intermediate space between these extremes. On the other hand, the [O/Fe] distribution from the exponential model is the most strongly peaked at the Solar value, while the plateau ($W=1$ Gyr) and triple-system models produce nearly identical distributions in the low-$\alpha$ regime. The exponential ($\tau=3$ Gyr) and plateau ($W=0.3$ Gyr) models, while not shown, produce similar abundance tracks to the plateau ($W=1$ Gyr) and exponential ($\tau=1.5$ Gyr) models, respectively.

@@

While trends with $R_{\rm gal}$ and $|z|$ are similar, the shape of the distribution varies greatly with the chosen SFH. The inside-out and late-burst models produce similar distributions because of the similarity of their underlying SFHs \textemdash both skew heavily toward near-Solar [O/Fe], although the late-burst model produces a slightly broader peak and a less-prominent high-[O/Fe] tail. The only region which shows any significant skew toward high [O/Fe] is $R_{\rm gal}=3-5$ kpc and $|z|=1-2$ kpc, but the shift to higher [O/Fe] at high latitudes is gradual and does not produce the notable trough at ${\rm [O/Fe]}\approx0.2$ which is seen in the APOGEE data.

@@

The two-infall SFH model produces \textit{three} distinct modes at ${\rm [O/Fe]}\approx 0.0$, $0.2$, and $0.4$. At small $R_{\rm gal}$ and with increasing $|z|$, the low-$\alpha$ peak decreases in prominence and the high-$\alpha$ peak increases, but the intermediate peak is a striking feature at all latitudes which is at odds with the APOGEE data.

@@

Figure \ref{fig:ofe-df-dtd} shows [O/Fe] distributions produced by simulations with the same SFH but a range of different DTDs. We choose the early-burst SFH because it produces distinct high- and low-$\alpha$ sequences, which makes the effect of the DTD on the former more apparent. The most apparent effect of the DTD is to shift the mode of the high-$\alpha$ sequence. The two-population DTD, which has the most prompt SNe Ia of any model in this paper, places the high-$\alpha$ sequence at ${\rm [O/Fe]}\approx 0.2$, while the triple-system DTD, which has the fewest prompt SNe Ia, places it $\sim0.2$ dex higher at ${\rm [O/Fe]}\approx0.4$. This metric alone would seem to favor the exponential ($\tau=1.5$ Gyr) or plateau ($W=0.3$ Gyr; not shown) models; however,

@@

The inside-out and late-burst models produce similar results, the main difference being a slightly broader core of the abundance distribution around ${\rm [O/Fe]}\approx0.05$ dex. The Gaussian burst in SFH in the late-burst model introduces a loop in the ISM abundance track, where an uptick in star formation at $t\approx11$ Gyr raises the CCSN rate, leading to a slight increase in [O/Fe] before the subsequent increase in the SN Ia rate brings it back down. This same pattern is seen much more strongly in the abundance tracks for the two-infall model: here, the significant infall of pristine gas at $t=4$ Gyr leads to rapid dilution of the ISM, followed by a large burst in the SFR which raises [O/Fe] by $\sim 0.2$ dex. In this case the second infall produces a ``ridge line'' in the abundance distribution at ${\rm [O/Fe]}\approx 0.2$. Finally, the early-burst model shows the clearest separation between a high-[O/Fe] track which follows the ISM abundance from low metallicities, and a low-[O/Fe] sequence composed of stars with a wide range of birth radii. The early-burst model sees a higher density of points which closely follow the ISM track at ${\rm [Fe/H]}\lesssim 0.5$ which is likely the result of an increased SFR at early times relative to the other models.

@@

For the early-burst SFH model, the simulated stellar ages are almost perfectly aligned with the data for ${\rm [O/Fe]}\gtrsim 0.2$. The rapid rise in the SFE at early times delays the descent to lower [O/Fe] values and produces a clump of low-metallicity, high-[O/Fe] stars at an age of $\sim10$ Gyr. The early-burst model produces a sharp transition between a linear relationship between [O/Fe] and age for stars older than $\sim5$ Gyr and a flat relationship for the younger stars, and it too misses the lowest [O/Fe] bin in the data. Finally, the two-infall SFH model produces the closest overall agreement between the simulated and observed median ages. Stars with ${\rm [O/Fe]}\gtrsim 0.25$ were produced in the first infall, while the second infall produces a clump of stars with similar metallicity, ages of $\sim8$ Gyr, and ${\rm [O/Fe]}\approx0.2$. There is an interesting small population of very old, low-$\alpha$ stars which is not seen in any of the other models. The two-infall model is the only one which plateaus at sub-Solar [O/Fe] for the youngest stars, leading to a very good match to the observed data at low [O/Fe].
